\textit{Supplementary materials for}: Text embedding models yield high resolution insights into conceptual knowledge from short multiple choice quizzes

### 2023-01-31 paper/main.tex
checking in changes before i go to a talk (everything is a mess!!!)

% supplemental figures
\newcommand{\individualKnowledgeMapsA}{S1}
\newcommand{\individualKnowledgeMapsB}{S2}
\newcommand{\individualKnowledgeMapsC}{S3}
\newcommand{\individualLearningMapsA}{S4}
\newcommand{\individualLearningMapsB}{S5}

@@

\title{Text embedding models reveal high resolution insights into conceptual
knowledge from short multiple choice quizzes}

@@

The difference between ``understanding'' and ``memorizing,'' as framed by
researchers in education, cognitive psychology, and cognitive
neuroscience~\citep[e.g.,][]{Kato40, Gall00, ScotEtal07, HallGree08, Macl05} has
profound analogs in the fields of natural language processing and natural
language understanding. For example, considering the raw contents of a document
(e.g., its constituent symbols, letters, and words) might provide some
information about what the document is about, just as memorizing a passage
might provide some ability to answer simple questions about it~\citep[e.g., whether
it contains words related to furniture versus physics;][]{LandDuma97,
BleiEtal03, BleiLaff06}. However, modern natural language processing
models~\citep[e.g.,][]{MikoEtal13a, CerEtal18, BrowEtal20} also attempt to
capture the deeper meaning \textit{underlying} those atomic elements. These
models consider not only the co-occurrences of those elements within and across
documents, but also patterns in how those elements appear across different
scales (e.g., sentences, paragraphs, chapters, etc.), the temporal and
grammatical properties of the elements, and other high-level characteristics of
how they are used~\citep{Mann20, Mann21a}. According to these models, the deep
conceptual meaning of a document may be captured by a feature vector in a
high-dimensional representation space, where nearby vectors reflect
conceptually related documents. A model that succeeds at capturing an analog of
``understanding'' is able to assign nearby feature vectors to two conceptually
related documents, \textit{even when the words contained in those documents
have very little overlap}.

@@

At its core, our main modeling approach is based around a simple assumption
that we sought to test empirically: all else being equal, knowledge about a
given concept is predictive of knowledge about similar or related concepts.
From a geometric perspective, this assumption implies that knowledge is
fundamentally ``smooth.'' In other words, as one moves through a space
representing an individual's knowledge (where similar concepts occupy nearby
coordinates), their ``level of knowledge'' should change relatively gradually
throughout that space. To begin to test this smoothness assumption, we sought
to track participants' knowledge and how it changed over time in response
to training.

@@

We asked participants in our study to complete brief multiple-choice quizzes
before, between, and after watching two lecture videos from the Khan
Academy~\citep{Khan04} platform (Fig.~\ref{fig:experiment}). The first lecture
video, entitled \textit{Four Fundamental Forces}, discussed the four
fundamental forces in physics: gravity, strong and weak interactions, and
electromagnetism. The second, entitled \textit{Birth of Stars}, provided an
overview of our current understanding of how stars form. We selected these
lessons to be (a) accessible to a broad audience, i.e., requiring minimal
prerequisite knowledge to understand; (b) conceptually related to each other,
i.e., covering at least \textit{some} similar or overlapping content; and (c)
largely independent of each other, i.e., focused on sufficiently different
material that understanding one did not require having seen the other. The two
videos we selected are introductory lectures that both belong to Khan Academy's
``Cosmology and Astronomy'' course domain, but are taken from different lecture
series (``Scale of the Universe'' and ``Stars, Black Holes, and Galaxies'' for
the first and second lectures, respectively).

We created a pool of multiple-choice quiz questions that would enable us to
evaluate participants' knowledge about each individual lecture along with
related content not specifically presented in either video (see
Tab.~\questions). Participants answered questions randomly drawn from each
content area (lecture 1, lecture 2, and general physics knowledge) on each of
the three quizzes. Quiz 1 was intended to assess participants' ``baseline''
knowledge before training, quiz 2 assessed knowledge after watching the
\textit{Four Fundamental Forces} video (i.e., lecture 1), and quiz 3 assessed
knowledge after watching the \textit{Birth of Stars} video (i.e., lecture 2).

@@

To study how participants' conceptual knowledge changed over the course of the
experiment, we first sought to characterize the abstract concepts presented to
them in each of the two lectures. We adapted an approach we developed in prior
work~\citep{HeusEtal21} to extract the latent themes from the lectures'
contents using a topic model~\citep{BleiEtal03}. Briefly, topic models take as
input a collection of text documents and learn a set of ``topics'' (i.e.,
latent themes) from their contents. Once fit, a topic model can be used to
transform arbitrary (potentially new) documents into sets of ``topic
proportions,'' describing the weighted blend of learned topics reflected in
their texts. We parsed automatically generated transcripts of the two lectures
into overlapping sliding windows, each containing the text of the lecture
transcript from a particular time range. We treated the set of text (across all
of these windows) as documents to fit our model
(Fig.~\ref{fig:sliding-windows}A; see \nameref{subsec:topic-modeling}).
Transforming the text from every window with our model yielded a
number-of-windows by number-of-topics (15) topic-proportions matrix that
described the unique mixture of broad themes from both lectures reflected in
each window's content. Intuitively, each window's ``topic vector'' (i.e.,
column of the topic-proportions matrix) is analogous to a coordinate in a
15-dimensional space whose axes are topics discovered by the model. Within this
space, each lecture's sequence of topic vectors (i.e., corresponding to sliding
windows parsed from its transcript) forms a \textit{trajectory} that captures
how its conceptual content unfolds over time (Fig.~\ref{fig:sliding-windows}B).
We resampled these trajectories to a resolution of one topic vector for each
second of video (i.e., 1~Hz).

Next, we sought to characterize what information participants' performance on
each of the three quizzes could provide about their conceptual knowledge at
that point in time. Traditional approaches to evaluating students'
performance on short-form knowledge assessments, such as those in our study,
entail computing the proportion of correctly answered questions to assign a
simple numeric score. While this may afford some measure of the
\textit{extent} of a learner's knowledge, such a score provides little
information to either the teacher or the learner about the particular
\textit{contents} of their knowledge---in other words, raw
``proportion-correct'' measures may capture \textit{how much} a student
knows, but not \textit{what} they know. For instance, suppose a participant
in our study was highly knowledgeable about fundamental physical forces (i.e
., lecture 1 content) but unfamiliar with how stars are formed (i.e., lecture
2 content), while a second participant was unfamiliar with fundamental forces
but had extensive prior knowledge about star formation. Since quiz 1
(completed before viewing either lecture) contained an equal number of
questions about each lecture's content (see Fig.~\ref{fig:experiment}), these
two participants could easily achieve the same proportion-correct score
despite their conceptual knowledge differing substantially. How might we
distinguish these individuals?

We hypothesized that our text embedding model, which we had trained on
transcripts of the two lectures to characterize their conceptual contents,
should also allow us to capture the conceptual knowledge probed by each quiz
question. If our model successfully represented information about the deeper
conceptual content of the lectures (i.e., beyond surface-level details such as
particular word choices) then we should be able to recover a correspondence
between their embeddings and embeddings of other text that reflects related
concepts (e.g., quiz questions that were ostensibly, by design, ``about'' one
of the two lecture) despite differences in exact verbiage. Intuitively, while
the content of each given lecture will naturally vary from moment to moment,
the particular combination of dimensions along which its trajectory varies
across moments correspond to topics that encode meaningful
``information''~\citep{Fish22} about that lecture's dynamic conceptual content.
We therefore expected that quiz questions pertaining to those same concepts
might express the same set of topics.

@@

We used our model to transform the text of each question in our pool into the
same embedding space as the two lectures' trajectories
(Fig.~\ref{fig:sliding-windows}C). This yielded a single 15-dimensional
coordinate (i.e., topic vector) for each question. We then computed the
variance of each topic's weight across timepoints of each lecture, and across
questions from each category (i.e., lecture 1-related, lecture 2-related, and
general physics knowledge; Fig.~\ref{fig:topics}A). Visual inspection of
Figure~\ref{fig:topics}A suggests a strong correspondence between the sets of
topics expressed by each lecture and its related questions, with only limited
overlap in the topics expressed by non-matched lecture-question set pairs. To
quantify these apparent similarities and differences, we computed the
correlation between each pair of topic-weight variance distributions
(Fig.~\ref{fig:topics}B). We found that our model represented the contents of
lecture-related questions using a combination of topic dimensions (and relative
variation along those topic dimensions) that was highly similar to that of
their reference lecture (\textit{Four Fundamental Forces} (\textit{FFF})
questions vs.~lecture: Pearson's $r(13) =
0.824,~p<0.001,~95\%~\mathrm{confidence~interval~(CI)} = [0.696, 0.973]$;
\textit{Birth of Stars} (\textit{BoS}) questions vs.~lecture: $r(13) =
0.801,~p<0.001,~95\%~\mathrm{CI} = [0.539, 0.958]$) but diverged significantly
from that of the non-reference lecture (\textit{FFF} questions vs.~\textit{BoS}
lecture: $r(13) = -0.620,~p = 0.014,~95\%~\mathrm{CI} = [-0.871, -0.326]$;
\textit{BoS} questions vs.~\textit{FFF} lecture: $r(13) = -0.550,~p =
0.034,~95\%~\mathrm{CI} = [-0.803, -0.246]$). This indicated that our model
captured the conceptual contents of the lectures and quiz questions
sufficiently well to differentiate between questions relating to one lecture
versus the other.

>>>>

% supplemental figures
\newcommand{\topicWeights}{S1} 
\newcommand{\individualKnowledgeMapsA}{S2}
\newcommand{\individualKnowledgeMapsB}{S3}
\newcommand{\individualKnowledgeMapsC}{S4}
\newcommand{\individualLearningMapsA}{S5}
\newcommand{\individualLearningMapsB}{S6}

@@

\title{Text embedding models yield high resolution insights into conceptual
knowledge from short multiple choice quizzes}

@@

The difference between ``understanding'' and ``memorizing,'' as framed by
researchers in education, cognitive psychology, and cognitive
neuroscience~\citep[e.g.,][]{Kato40, Gall00, ScotEtal07, HallGree08, Macl05}
has profound analogs in the fields of natural language processing and natural
language understanding. For example, considering the raw contents of a document
(e.g., its constituent symbols, letters, and words) might provide some
information about what the document is about, just as memorizing a passage
might provide some ability to answer simple questions about it. However, text
embedding models~\citep[e.g.,][]{LandDuma97, DeerEtal90, BleiEtal03,
BleiLaff06, MikoEtal13a, CerEtal18, BrowEtal20} also attempt to capture the
deeper meaning \textit{underlying} those atomic elements. These models consider
not only the co-occurrences of those elements within and across documents, but
also patterns in how those elements appear across different scales (e.g.,
sentences, paragraphs, chapters, etc.), the temporal and grammatical properties
of the elements, and other high-level characteristics of how they are
used~\citep{Mann20, Mann21a}. According to these models, the deep conceptual
meaning of a document may be captured by a feature vector in a high-dimensional
representation space, where nearby vectors reflect conceptually related
documents. A model that succeeds at capturing an analog of ``understanding'' is
able to assign nearby feature vectors to two conceptually related documents,
\textit{even when the words contained in those documents have very little
overlap}.

@@

At its core, our main modeling approach is based around a simple assumption
that we sought to test empirically: all else being equal, knowledge about a
given concept is predictive of knowledge about similar or related concepts.
From a geometric perspective, this assumption implies that knowledge is
fundamentally ``smooth.'' In other words, as one moves through a space
representing an individual's knowledge (where similar concepts occupy nearby
coordinates), their ``level of knowledge'' should change relatively gradually
throughout that space. To begin to test this smoothness assumption, we sought
to track participants' knowledge and how it changed over time in response to
training. Two overarching goals guide our approach. First, we want to gain
detailed insights into what learners know, at different points in their
training. For example, rather than simply reporting on the proportions of
questions participants answer correctly (i.e., their overall performance), we
seek estimates of their knowledge about a variety of specific concepts.
Second, we want our approach to be potentially scalable to large numbers of
concepts, courses, and students. This requires the conceptual content of
interest to be discovered \textit{automatically}, rather than relying on
manually produced ratings or labels.

@@

We asked participants in our study to complete brief multiple-choice quizzes
before, between, and after watching two lecture videos from the Khan
Academy~\citep{Khan04} platform (Fig.~\ref{fig:experiment}). The first lecture
video, entitled \textit{Four Fundamental Forces}, discussed the four
fundamental forces in physics: gravity, strong and weak interactions, and
electromagnetism. The second, entitled \textit{Birth of Stars}, provided an
overview of our current understanding of how stars form. We selected these
particular lectures to satisfy three general criteria. First, we wanted both
lectures to be accessible to a broad audience (i.e., with minimal prerequisite
knowledge) so as to limit the impact of prior training on our participants'
abilities to learn from the lectures. To this end, we selected two introductory
videos that were intended to be viewed at the start of students' training in
their respective content areas. Second, we wanted both lectures to have some
related content, so that we could test our approach's ability to distinguish
similar conceptual content. To this end, we chose two videos from the same (per
instructor annotations) Khan Academy course domain, ``Cosmology and
Astronomy.'' Third, we sought to minimize dependencies and specific overalp
between the videos. For example, we did not want participants' abilities to
understand one video to (directly) influence their abilities to understand the
other. To satisfy this last criterion, we chose videos from two different
lecture series (lectures 1 and 2 were from the ``Scale of the Universe'' and
``Stars, Black Holes, and Galaxies'' series, respectively).

We also wrote a set of multiple-choice quiz questions that we hoped would
enable us to evaluate participants' knowledge about each individual lecture,
along with related knowledge about physics not specifically presented in either
video (see Tab.~\questions~for the full list of questions in our stimulus
pool). Participants answered questions randomly drawn from each content area
(lecture 1, lecture 2, and general physics knowledge) on each of the three
quizzes. Quiz 1 was intended to assess participants' ``baseline'' knowledge
before training, quiz 2 assessed knowledge after watching the \textit{Four
Fundamental Forces} video (i.e., lecture 1), and quiz 3 assessed knowledge
after watching the \textit{Birth of Stars} video (i.e., lecture 2).

@@

To study in detail how participants' conceptual knowledge changed over the
course of the experiment, we first sought to model the conceptual content
presented to them at each moment throughout each of the two lectures. We
adapted an approach we developed in prior work~\citep{HeusEtal21} to identify
the latent themes in the lectures using a topic model~\citep{BleiEtal03}.
Briefly, topic models take as input a collection of text documents and learn a
set of ``topics'' (i.e., latent themes) from their contents. Once fit, a topic
model can be used to transform arbitrary (potentially new) documents into sets
of ``topic proportions,'' describing the weighted blend of learned topics
reflected in their texts. We parsed automatically generated transcripts of the
two lectures into overlapping sliding windows, where each window contained the
text of the lecture transcript from a particular time range. We treated the set
of text snippets (across all of these windows) as documents to fit our model
(Fig.~\ref{fig:sliding-windows}A; see \nameref{subsec:topic-modeling}).
Transforming the text from every sliding window with our model yielded a
number-of-windows by number-of-topics (15) topic-proportions matrix that
described the unique mixture of broad themes from both lectures reflected in
each window's content. Each window's ``topic vector'' (i.e., column of the
topic-proportions matrix) is a coordinate in a 15-dimensional space whose axes
are topics discovered by the model. Within this space, each lecture's sequence
of topic vectors (i.e., corresponding to its transcript's overlapping text
snippets across sliding windows) forms a \textit{trajectory} that captures how
its conceptual content unfolds over time (Fig.~\ref{fig:sliding-windows}B). We
resampled these trajectories to a resolution of one topic vector for each
second of video (i.e., 1~Hz).

We hypothesized that a topic model trained on transcripts of the two lectures,
should also capture the conceptual knowledge probed by each quiz question. If
indeed the topic model could capture information about the deeper conceptual
content of the lectures (i.e., beyond surface-level details such as particular
word choices) then we should be able to recover a correspondence between each
lecture and questions \textit{about} each lecture. Importantly, such a
correspondence could not solely arise from superficial text matching between
lecture transcripts and questions, since the lectures and questions used
different words. Simply comparing the average topic weights from each lecture
and question sets (averaging across time and questions, respectively) reveals a
striking correspondence (Fig.~\topicWeights). Specifically, the average topic
weights from lecture 1 are strongly correlated with the average topic weights
from lecture 1 questions ($r(13) = XX, p = XX$, 95\% confidence interval
(CI)~$= XX$), and the average topic weights from lecture 2 are strongly
correlated with the average topic weights from lecture 2 questions ($r(13) =
XX, p = XX, CI = XX$). At the same time, the average topics from two lectures
are \textit{negatively} correlated ($r(13) = XX, p = XX, CI = XX$). The full
set of pairwise comparisons between topic vectors for the lectures and each
question set is reported in Figure~\topicWeights.

@@

Another, more sensitive, way of summarizing the conceptual content of the
lectures and questions is to look at \textit{variability} in how topics are
weighted over time and across different questions (Fig.~\ref{fig:topics}).
Intuitively, the variability in the expression of a given topic relates to how
much ``information''~\citep{Fish22} the lecture (or questions) reflect about
that topic. For example, suppose a given topic is weighted on heavily
throughout a lecture. That topic might be characteristic of some aspect or
property of the lecture \textit{overall} (conceptual or otherwise), but unless
the topic's weights changed in meaningful ways over time, the topic would be a
poor indicator of any \textit{specific} conceptual content in the lecture. We
therefore also compared the variance in topic weights (across time or
questions) between the lectures and questions.  [ADD BACK CORRELATION RESULTS]

\textbf{JRM STOPPED HERE...}

We used the fitted topic model to transform the text of each question in our
stimulus pool into the same embedding space as the two lectures' trajectories
(Fig.~\ref{fig:sliding-windows}C). This yielded a single 15-dimensional
coordinate (i.e., topic vector) for each question. We then computed the
variance of each topic's weight across timepoints of each lecture, and across
questions from each category (i.e., questions about lecture 1, lecture 2, and
general physics knowledge; Fig.~\ref{fig:topics}A). Visual inspection of
Figure~\ref{fig:topics}A suggests a strong correspondence between the sets of
topics expressed by each lecture and its related questions, with only limited
overlap in the topics expressed by non-matched lecture-question set pairs. To
quantify these apparent similarities and differences, we computed the
correlations between each pair of topic-weight variance vectors (i.e., vectors
containing the variance across timepoints or questions, for each topic;
Fig.~\ref{fig:topics}B).

We found that our model represented the contents of
lecture-related questions using a combination of topic dimensions (and relative
variation along those topic dimensions) that was highly similar to that of
their reference lecture (\textit{Four Fundamental Forces} (\textit{FFF})
questions vs.~lecture: Pearson's $r(13) =
0.824,~p<0.001,~95\%~\mathrm{confidence~interval~(CI)} = [0.696, 0.973]$;
\textit{Birth of Stars} (\textit{BoS}) questions vs.~lecture: $r(13) =
0.801,~p<0.001,~95\%~\mathrm{CI} = [0.539, 0.958]$) but diverged significantly
from that of the non-reference lecture (\textit{FFF} questions vs.~\textit{BoS}
lecture: $r(13) = -0.620,~p = 0.014,~95\%~\mathrm{CI} = [-0.871, -0.326]$;
\textit{BoS} questions vs.~\textit{FFF} lecture: $r(13) = -0.550,~p =
0.034,~95\%~\mathrm{CI} = [-0.803, -0.246]$). This indicated that our model
captured the conceptual contents of the lectures and quiz questions
sufficiently well to differentiate between questions relating to one lecture
versus the other.

### 2023-01-31 paper/supplement.tex
checking in changes before i go to a talk (everything is a mess!!!)

\title{\textit{Supplementary materials for}: Text embedding models reveal high
resolution insights into conceptual knowledge from short multiple choice
quizzes}

@@

@@

    \label{tab:topics}

\end{tabular}
\end{tiny}
\caption{\textbf{Topics.} The table displays the 10 top-weighted word from
    each of the 15 topics identified from sliding windows of two course video
    transcripts (see \textit{Constructing text embeddings of multiple videos
    and questions}).}
\end{table}

>>>>

\title{\textit{Supplementary materials for}: Text embedding models yield high
resolution insights into conceptual knowledge from short multiple choice
quizzes}

@@

\newcommand{\topics}{3}

@@

\end{tabular}
\end{tiny}

\caption{\textbf{Topics.} The table displays the 10 top-weighted word from
    each of the 15 topics identified from sliding windows of two course video
    transcripts (see \textit{Constructing text embeddings of multiple videos
    and questions}).}
    \label{tab:topics}
\end{table}

\begin{figure}[tp]
    \centering
    \includegraphics[width=\textwidth]{figs/topic-weights}

\caption{\textbf{Averarge topic weights for each lecture and question type}.
The bar plots display the variance of each topic's weight across lecture
timepoints (top row) and questions (bottom row); colors denote topics. The
top-weighted words from the most heavily weighted topics are displayed in the
upper right (orange: topic 2; yellow-green: topic 4). The top-weighted words
from the full set of topics may be found in Table~S2. Also see
Figure~\topics A in the main text.}

    \label{fig:topics}
\end{figure}
